Supplementary Information for RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2014-03-13 doc/bioinformatics2014/document.tex
Update manuscript

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}

@@

For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$. The model generative process is as follows:

@@

@@

@@

Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF. Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations. Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample. In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T. Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the figure.
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}} - \mu^{\text{control}} \geq \tau$.}
\label{fig:brca_MAF}
\end{center}
\end{figure}
%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
%TODO: add star to test that have mudelta > .16 Done.
%TODO: set tau=0.16 for mudelta test.  Done

>>>>

\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}

@@

For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$. $ J $ is the length of gene sequence and $ N $ is the number of technical replicate gene sequence in the sample.  \label{replicate definition} Technical replicates are used to establish experimental variability in Next Generation Sequencing procedure~\citep{quackenbush2002microarray,robasky2013role}, even though multiple replicates are not indispensable for RVD2.

The model generative process is as follows:

@@

\label{Improvement of RVD2}
The improvement of RVD2 upon RVD firstly lies in the $ \text{Beta}(\mu_0, M_0) $ prior on local error rate $ \mu_j $, which accurately captured the global across-position error rate distribution. The prior distribution also  allows to $ \mu_j $ borrow information from adjacent positions and prevent RVD2 from failure when the read depth is too low to bring out none-zero $ r_{ji} $. Besides the improvement on model structure, RVD2 developed a novel Bayesian Hypothesis Testing algorithm comparing to a frequentist normal z-test in RVD. The performance comparison between RVD and RVD2 will be elaborated in Section~\ref{sec:comparison}.

@@

\begin{figure*}[!tpb]
\begin{center}
\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$  is significantly different from $ \mu^{\text{control}}$.}
\label{fig:brca_MAF}
\end{center}
\end{figure*}
%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
%TODO: add star to test that have mudelta > .16 Done.
%TODO: set tau=0.16 for mudelta test.  Done

@@

Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF. Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations. Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample. In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, \\ chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T. Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the figure.
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

### 2014-03-13 doc/bioinformatics2014/supplementary_info.tex
update manuscript in bioinformatics2014
Update manuscript

%%%%%%%%%%%%%%%%%%%
% Appendix A: Parameter Initialization
%%%%%%%%%%%%%%%%%%%
\section{Parameter Initialization}\label{sec:appendix_mom}
Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$. Therefore the MoM estimator is
\begin{equation}
	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
\end{equation}

We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model. The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$. The first and second population moments are
\begin{eqnarray}
	E[\theta_{ji}] =& \mu_j,\\
	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
\end{eqnarray}
The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$. Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
\begin{eqnarray}
	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
\end{eqnarray}

Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
\begin{eqnarray}
	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
\end{eqnarray}

@@

>>>>

%%%%%%%%%%%%%%%%%%%
% Appendix A: Parameter Initialization
%%%%%%%%%%%%%%%%%%%
\section{Parameter Initialization}\label{sec:appendix_mom}
Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$. Therefore the MoM estimator is
\begin{equation}
	\hat{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
\end{equation}

We take the MoM estimate, $\hat{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model. The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$. The first and second population moments are
\begin{eqnarray}
	E[\theta_{ji}] =& \mu_j,\\
	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
\end{eqnarray}
The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$. Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
\begin{eqnarray}
	\hat{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}, \\
	\hat{M_j} =& \frac{ \hat{\mu}_j (1 - \hat{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}^2 } -1.
\end{eqnarray}

Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
\begin{eqnarray}
	\hat{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \hat{\mu}_j \\
	\hat{M}_0 =& \frac{ \hat{\mu}_0 (1 - \hat{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \hat{\mu}_j^2 } -1.
\end{eqnarray}

@@

\section{Histogram of $ \mu_j $ for mutations detected by RVD2 in HCC1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:hist_neg}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:hist_pos}
\end{center}
\end{figure}

\section{Comparison between RVD2, VarScan2 somatic and muTect in Clincal HCC1187 dataset}
\begin{figure}[H]
\begin{center}
\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_RVD2.pdf}
\caption{RVD2.}
\label{fig:rd_RVD2}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_mutect.pdf}
\caption{muTect.}
\label{fig:rd_mutect}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_true_positives.pdf}
\caption{HCC1187 varscan2 true positives.}
\label{fig:rd_varscan2_TP}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_negatives.pdf}
\caption{ HCC1187 varscan2 false negatives.}
\label{fig:rd_varscan2_FN}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_positives.pdf}
\caption{HCC1187 varscan2 false positives}
\label{fig:rd_varscan2_FP}
\end{center}
\end{figure}
